Temporal semantics for a live coding language

### 2014-05-08 paper/paper.tex
some work on monad laws and warnings

The \ensuremath{\Conid{Temporal}} moand is ``weak'', in the sense that the standard monad
laws do not hold in general.  For example, one of the unit laws is that:
%%
\begin{equation}
%|(return x) >>= (\y -> f y)| \equiv |f x|
m >>= return \equiv m
\label{law-example}
\end{equation}
%%
where \ensuremath{\Varid{m}\mathbin{::}\Conid{Temporal}\;\Varid{a}}.  However, for \ensuremath{\Conid{Temporal}} this law is
violated in cases where \ensuremath{\Varid{m}} depends on the current time. For example, let
\ensuremath{\Varid{m}} be defined:
%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{m}\mathrel{=}\mathbf{do}\;{}\<[10]%
\>[10]{}\Varid{kernelSleep}\;\mathrm{1}{}\<[E]%
\\
\>[10]{}\Varid{start}\leftarrow \Varid{start}{}\<[E]%
\\
\>[10]{}\Varid{end}\leftarrow \Varid{time}{}\<[E]%
\\
\>[10]{}\Varid{return}\;(\Varid{diffTime}\;\Varid{end}\;\Varid{start})\mbox{\onelinecomment  duration of computation}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%%
Then we can run an experiment in GHCi to see that different results are possible:
%%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbin{*}\Conid{Main}\mathbin{>}\Varid{runTime}\mathbin{\$}\Varid{m}\bind \Varid{return}{}\<[E]%
\\
\>[B]{}\mathrm{1.001113}\;\Varid{s}{}\<[E]%
\\
\>[B]{}\mathbin{*}\Conid{Main}\mathbin{>}\Varid{runTime}\mathbin{\$}\Varid{m}{}\<[E]%
\\
\>[B]{}\mathrm{1.00114}\;\Varid{s}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% The difference in results follows from the additional reduction
required on \ensuremath{(\bind )} in the first case (left-hand side of
\eqref{law-example}).

>>>>

The \ensuremath{\Conid{Temporal}} moand is ``weak'', in the sense that the standard monad
laws do not hold in general.  For example, one of the unit laws is that:
%%
\begin{equation}
%|(return x) >>= (\y -> f y)| \equiv |f x|
m >>= return \equiv m
\label{law-example}
\end{equation}
%%
where \ensuremath{\Varid{m}\mathbin{::}\Conid{Temporal}\;\Varid{a}}.  However, for \ensuremath{\Conid{Temporal}} this law is
violated in cases where \ensuremath{\Varid{m}} depends on the current time. For example, let
\ensuremath{\Varid{m}} be defined:
%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{m}\mathrel{=}\mathbf{do}\;{}\<[10]%
\>[10]{}\Varid{kernelSleep}\;\mathrm{1}{}\<[E]%
\\
\>[10]{}\Varid{start}\leftarrow \Varid{start}{}\<[E]%
\\
\>[10]{}\Varid{end}\leftarrow \Varid{time}{}\<[E]%
\\
\>[10]{}\Varid{return}\;(\Varid{diffTime}\;\Varid{end}\;\Varid{start})\mbox{\onelinecomment  duration of computation}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%%
Then we can run an experiment in GHCi to see that different results are possible:
%%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbin{*}\Conid{Main}\mathbin{>}\Varid{runTime}\mathbin{\$}\Varid{m}\bind \Varid{return}{}\<[E]%
\\
\>[B]{}\mathrm{1.001113}\;\Varid{s}{}\<[E]%
\\
\>[B]{}\mathbin{*}\Conid{Main}\mathbin{>}\Varid{runTime}\mathbin{\$}\Varid{m}{}\<[E]%
\\
\>[B]{}\mathrm{1.00114}\;\Varid{s}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
(note, these results are also non-deterministic).  The difference in
results follows from the additional reduction required on \ensuremath{(\bind )} in
the first case (left-hand side of \eqref{law-example}). Note that we
calculate a duration here since using the absolute time produced by
\ensuremath{\Varid{time}} would be disingenuous, since we are evaluating \ensuremath{\Varid{m}\bind \Varid{return}}
and \ensuremath{\Varid{m}} at different start times.

\paragraph{Laws with respect to $\epsilon$}
